---
title: "Denver Crime Visualization"
author: "cyorir"
date: "September 18, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Denver Crime Data

The city of Denver provides data on various types of crime in the city. As of September 18, 2018, this includes datasets for crime, hate crime, and marijuana-related crime. In addition, data on police pedestrian stops is provided. These datasets can be used to answer some interesting questions, such as which neighborhoods suffer from specific types of crime, or which types of crime might occur in the same location. These datasets can be accessed at the URLs listed below.

https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-crime

https://www.denvergov.org/opendata/dataset/hate-crimes

https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-crime-marijuana

https://www.denvergov.org/opendata/dataset/city-and-county-of-denver-police-pedestrian-stops-and-vehicle-stops

## Denver Hate Crime Dataset

The hate crimes data set provides information about hate crimes recorded by the Denver Police Department. This includes information about the location of the crime and the type of hate crime. The data indicate the relative frequency of different types of hate crime in Denver.

```{r bias_types}
data <- read.csv("BiasMotivatedCrimes.csv")
summary(data["Bias.Type"])
```
Other summaries might be interesting, including location description and offense description.

```{r loc_off}
print(summary(data["OFFENSE_DESC"]))
summary(data["Location_Description"])
```

The data can be mapped to show where in Denver hate crimes occur. Since the X and Y coordinates are in a state plane coordinate system, they must first be mapped to latitude and longitude so that they can be plotted.

```{r hc_transform}
library("rgdal")
transformCoords <- function(xy) {
  colnames(xy) <- c("x","y")
  x <- xy["x"][!is.na(xy["x"])]
  #note that for this dataset either both x and y
  #are present or both are missing
  y <- xy["y"][!is.na(xy["x"])]
  xy <- data.frame(cbind(x,y))
  xy <- SpatialPoints(xy, CRS("+init=EPSG:3502"))
  proj4string <- "+init=epsg:4326"
  pj <- spTransform(xy,proj4string)
  latlon <- data.frame(lat=pj$y, lon=pj$x)
  latlon
}
xy <- data[c("X_COORDINATE", "Y_COORDINATE")]
head(transformCoords(xy))
```

Now that the approximate latitude and longitude are available for each crime, the crimes can be mapped over the city of Denver. 

```{r hc_map}
library(leaflet)
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Note that data for the suburbs are not included, and each marker represents a nearby block rather than the precise location of the crime. The data can be mapped for a specific type of offense. In this case, the data for hate graffiti is mapped.

```{r graffiti}
library(dplyr)
data_graffiti <- data %>% filter(OFFENSE_DESC == "CRIMINAL MISCHIEF - GRAFFITI")
xy <- data_graffiti[c("X_COORDINATE", "Y_COORDINATE")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

The data for different types of bias may also be mapped.

Map of anti-black hate crimes:

```{r anti-black}
data_black <- data %>% filter(Bias.Type == "ANTI-BLACK")
xy <- data_black[c("X_COORDINATE", "Y_COORDINATE")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Map of anti-white hate crimes:

```{r anti-white}
data_white <- data %>% filter(Bias.Type == "ANTI-WHITE")
xy <- data_white[c("X_COORDINATE", "Y_COORDINATE")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Map of anti-hispanic hate crimes:

```{r anti-hispanic}
data_hispanic <- data %>% filter(Bias.Type == "ANTI-HISPANIC")
xy <- data_hispanic[c("X_COORDINATE", "Y_COORDINATE")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Map of anti-Semitic hate crimes:

```{r anti-jewish}
data_jewish <- data %>% filter(Bias.Type == "ANTI-JEWISH")
xy <- data_jewish[c("X_COORDINATE", "Y_COORDINATE")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Note that each type of bias follows a different distribution. Also note that this only considers crimes attributed to a single type of bias.

Hopefully this provides an example of how Denver's open data can be used.

## Denver Marijuana Crime Dataset

Marijuana has been a hot topic in Colorado since Amendment 64 was passed, legalizing the recreational use of marijuana. Denver now provides open data about marijuana-related crimes. Crimes are separated into industry and non-industry crimes. These are explained as follows: "Industry-related crimes involve marijuana and licensed marijuana facilities. These reported crimes are committed against the licensed industry or by the industry itself. Non-Industry crimes are crimes reported where marijuana is the primary target in the commission of these crime but the marijuana has no readily apparent tie to a licensed operation." 

```{r marijuana}
data <- read.csv("crime_marijuana.csv")
summary(data["MJ_RELATION_TYPE"])
```

Data about the type of offense is also included. Burglary is by far the most common type of offense.

```{r mj_offenses}
summary(data["OFFENSE_CATEGORY_ID"])
```

Crimes can be grouped by neighborhood.

```{r mj_neighborhoods}
summary(data["NEIGHBORHOOD_ID"])
```

The dataset includes the coordinates of the crimes. As with the hate crimes dataset, the coordinates must be transformed from the state plane coordinate system to latitude and longitude. Marijuana-related burglaries are mapped.

```{r mj_burglary}
data_burglary <- data %>% filter(OFFENSE_CATEGORY_ID == "Burglary")
xy <- data_burglary[c("GEO_X", "GEO_Y")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

There is also the map of all other marijuana-related crimes.

```{r mj_other}
data_burglary <- data %>% filter(OFFENSE_CATEGORY_ID != "Burglary")
xy <- data_burglary[c("GEO_X", "GEO_Y")]
m <- leaflet(data=transformCoords(xy)) %>%
  addTiles() %>%
  addMarkers(~lon,~lat, clusterOptions = markerClusterOptions())
m
```

Note that the two maps have a similar distribution.

## Denver Crime Dataset

The Denver Open Data Catalog includes a dataset for crime in general. This dataset is much larger and more comprehensive than either the hate crime dataset or marijuana crime datasets.

```{r crime}
data <- read.csv("crime.csv")
summary(data["OFFENSE_CATEGORY_ID"])
```

The data includes the neighborhood in which each crime was committed.

```{r crime_neighborhood}
summary(data["NEIGHBORHOOD_ID"])
```

Note that Montbello and Five-Points appeared in the list of top locations for marijuana crime as well.

The data include a lot of traffic incidents which may not be considered crimes. These are filtered out below.

```{r crime_traffic}
print(nrow(data))
data <- data %>%
    filter(IS_CRIME == 1)
print(nrow(data))
```

The "OFFENSE_TYPE_ID" provides more detail on what crime was committed than the "OFFENSE_CATEGORY_ID". For example, the frequencies of different types of larceny can be viewed.

```{r crime_larceny}
data_larceny <- data %>% filter(OFFENSE_CATEGORY_ID == "larceny")
summary(data_larceny["OFFENSE_TYPE_ID"])
```

The dataset includes latitude and longitude, so no conversion between coordinate systems is necessary to map the data. Since the dataset is so large, only the first 10,000 valid crime entries will be mapped.

```{r crime_map}
xy <- data[c("GEO_LON", "GEO_LAT")]
xy <- xy %>% filter(!is.na(GEO_LON)) %>% filter(!is.na(GEO_LAT)) %>% filter(GEO_LON < -100)
xy <- head(xy, 10000)
m <- leaflet(data=xy) %>%
  addTiles() %>%
  addMarkers(~GEO_LON,~GEO_LAT, clusterOptions = markerClusterOptions())
m
```

Seeing the time difference between reporting of the crime and occurrence of the crime (in days) may be of interest.

```{r crime_dates}
library(lubridate)
data_timing <- data %>%
            mutate(REP_OCC_DIFF = as.duration(mdy_hms(REPORTED_DATE)-
                     mdy_hms(FIRST_OCCURRENCE_DATE))/ddays(1))
print(head(data_timing["REP_OCC_DIFF"]))
summary(data_timing["REP_OCC_DIFF"])
```

## Denver Police Stops Dataset

A fourth crime dataset provided by the Denver Open Data Catalog provides information on police stops of pedestrians and vehicles. This dataset is about as large as the crime dataset.

```{r police_stops}
data <- read.csv("police_pedestrian_stops_and_vehicle_stops.csv")
summary(data["PROBLEM"])
```

Once again, the neighborhoods can be viewed.

```{r police_stops_neighborhoods}
summary(data["NEIGHBORHOOD_NAME"])
```

Five-Points, Montbello, and Stapleton appear again.

The number of arrests made can be determined through the call_disposition.

```{r police_stops_arrests}
arrests <- data %>% filter(CALL_DISPOSITION == "Arrest Made")
nrow(arrests)
```

What period of time does the dataset cover?

```{r police_stops_dates}
data <- data %>% mutate(DATES = ymd_hms(TIME_PHONEPICKUP))
data <- data %>% arrange(DATES)
print(head(data["DATES"]))
print(tail(data["DATES"]))
```

Like the crime dataset, this dataset includes the latitude and longitude of incidents so no conversion from the state plane system is necessary. They aren't mapped here because the dataset is so large, but the process of mapping this data would be similar to mapping the crime dataset.

## Conclusion

Hopefully this report provides a decent example of how Denver's open crime data can be mapped. At present, it is still a challenge to integrate this data with similar crime data from the suburbs. This was only exploratory data analysis, hopefully it can provide a basis for more in depth analysis. The data from these datasets could be used together with population data to provide a view of per-capita crime.
